Discrete Quantum Theories and Computing

### 2018-03-21 dissertation.tex
dissertation.lyx "3.4 Discrete Quantum Theory (I)" Updating...

In this section we want to build a theory of discrete vector spaces
that approximates as closely as possible the features of conventional
quantum theory. Such a structure would ideally consist of the following:
(i) a vector space over the field of complex numbers, and (ii) an
inner product $\ip{\Phi}{\Psi}$ associating to each pair of vectors
a complex number, and satisfying the following properties: 
\begin{enumerate}[label=(\Alph{enumi})]
\item $\ip{\Phi}{\Psi}$ is the complex conjugate of $\ip{\Psi}{\Phi}$; 
\item $\ip{\Phi}{\Psi}$ is conjugate linear in its first argument and linear
in its second argument; 
\item $\ip{\Psi}{\Psi}$ is always non-negative and is equal to~0 only
if $\ket{\Psi}$ is the zero vector. 
\end{enumerate}
It turns out that a vector space defined over a finite field cannot
have an inner product satisfying the properties above. However, we
will introduce an Hermitian ``dot product'' satisfying some of those
properties.

@@

\begin{definition}[Hermitian dot product] Given vectors $\ket{\Phi}$
and $\ket{\Psi}\in{\cal H}$, the Hermitian dot product of these vectors
is: 
\begin{equation}
\ip{\Phi}{\Psi}=\sum_{i=0}^{D-1}\beta_{i}^{p}\alpha_{i}\,.
\end{equation}
\end{definition} Two vectors $\ket{\Phi}$ and $\ket{\Psi}\in{\cal H}$
are said to be orthogonal if $\ip{\Phi}{\Psi}=0$. This product satisfies
conditions A and B for inner products but violates condition C since
in every finite field there always exists a non-zero vector $\ket{\Psi}$
such that $\ip{\Psi}{\Psi}=0$. The reason is that addition in finite
fields eventually ``wraps around'' (because of their cyclic or modular
structure), allowing the sum of non-zero elements to be zero. The
fraction of non-zero vectors satisfying $\ip{\Psi}{\Psi}=0$ decreases
with the order $p$.

@@

>>>>

In this section we want to build a theory of discrete vector spaces
that approximates as closely as possible the features of conventional
quantum theory. Such a structure would ideally consist of the following:
(i) a vector space over the field of complex numbers, and (ii) an
inner product $\ip{\Phi}{\Psi}$ associating to each pair of vectors
a complex number, and satisfying the following properties: 
\begin{enumerate}[label=(\Alph{enumi})]
\item \label{enu:inner-product-complex-conjugate}$\ip{\Phi}{\Psi}$ is
the complex conjugate of $\ip{\Psi}{\Phi}$; 
\item \label{enu:inner-product-linear}$\ip{\Phi}{\Psi}$ is conjugate linear
in its first argument and linear in its second argument; 
\item \label{enu:inner-product-positive-definite}$\ip{\Psi}{\Psi}$ is
always non-negative and is equal to~0 only if $\ket{\Psi}$ is the
zero vector. 
\end{enumerate}
It turns out that a vector space defined over a finite field cannot
have an inner product satisfying the properties above. However, we
will introduce an Hermitian ``dot product'' satisfying some of those
properties.

@@

\begin{definition}[Hermitian dot product] Given vectors $\ket{\Phi}$
and $\ket{\Psi}\in{\cal H}$, it can be shown~\cite{grove2002classical}
the Hermitian dot product is always reducible to the form 
\begin{equation}
\ip{\Phi}{\Psi}=\sum_{i=0}^{D-1}\beta_{i}^{p}\alpha_{i}\,.
\end{equation}
\end{definition} Two vectors $\ket{\Phi}$ and $\ket{\Psi}\in{\cal H}$
are said to be orthogonal if $\ip{\Phi}{\Psi}=0$. This product satisfies
conditions \ref{enu:inner-product-complex-conjugate} and \ref{enu:inner-product-linear}
for inner products but violates condition \ref{enu:inner-product-positive-definite}
since in every finite field there always exists a non-zero vector
$\ket{\Psi}$ such that $\ip{\Psi}{\Psi}=0$. The reason is that addition
in finite fields eventually ``wraps around'' (because of their cyclic
or modular structure), allowing the sum of non-zero elements to be
zero. The fraction of non-zero vectors satisfying $\ip{\Psi}{\Psi}=0$
decreases with the order $p$.

@@

\section{Irreducible Discrete $D$-dimensional States: Generalized Discrete
Bloch Sphere\label{DQCnqubitBloch.sec}}

In the one-qubit state with coefficients in $\Fpp$, the discrete
analog of the Bloch sphere is constructed by exact analogy to the
continuous case: we first require that the coefficients of the single
qubit basis obey 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=1\label{DQA2.eq}
\end{equation}
in the discrete field. In \ref{appU.sec}, we show that there are
$p(p^{2}-1)$ such values.\todo{{Check reference...}} Given this
requirement, which is similar in form to the conservation of probability,
but not as useful due to the lack of orderable probability values,
we can immediately conclude that the discrete analog of the Hopf fibration
is again 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\label{DQA6.eq}\\
Z & = & \left|\alpha_{0}\right|^{2}-\left|\alpha_{1}\right|^{2}={x_{0}}^{2}+{y_{0}}^{2}-{x_{1}}^{2}-{y_{1}}^{2}\,.\nonumber 
\end{eqnarray}
but now with all computations in $\pmod{p}$. At this point one simply
writes down all possible discrete values for the complex numbers $\left(\alpha_{0},\alpha_{1}\right)$
satisfying Eq.~(\ref{DQA2.eq}) and enumerates those that project
to the same value of $\{X,Y,Z\}$. This equivalence class is the discrete
analog of the circle in the complex plane that was eliminated in the
continuous case. In \ref{appH.sec}, we show that $p+1$ discrete
values of $\{\alpha_{0},\alpha_{1}\}$ with unit norm map to the same
point under the Hopf map \Eq{DQA6.eq}; we may think of these as
discrete circles or projective lines of equivalent, physically indistinguishable,
complex phase. The surviving $p(p-1)$ values of $\{\alpha_{0},\alpha_{1}\}$
correspond to irreducible physical states of the discrete single qubit
system. %% Several examples are shown in Figure \ref{bloch.fig}. %%Thus,
for example, choosing the underlying field to be $\Fppx{3}$, there
are exactly 6 single-qubit state vectors to populate the Bloch sphere;
the four equivalent phase-multiples mapping to each of the six points
on the $\Fppx{3}$ Bloch sphere are collapsed and regarded as physically
indistinguishable. In Figure \ref{bloch.fig}, we plot the irreducible
states on the Bloch sphere for $p=3$, $7$, and $11$. Note that
the Cartesian lengths of the real vectors corresponding to the points
on the Bloch sphere vary considerably due to the nature of discrete
fields; we have artificially normalized them to a ``continuous world''
unit radius sphere for conceptual clarity.

\begin{figure}[htb]
%\vspace{-0.2in}  %% \includegraphics[width=0.9\columnwidth]{./Bloch_sphere.pdf}%%\includegraphics[width=2.0\columnwidth]{./picA.png}%%\includegraphics[width=2.0\columnwidth]{./picB.png}

\begin{centering}
\hspace*{-0.5cm}\includegraphics[width=1.1\columnwidth]{\string"../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/picC\string".png}
%%\includegraphics[width=0.5\textwidth]{./picAB.eps}\vspace{-0.2in}
 %% adjust blank space?
\par\end{centering}
\caption{Schematically normalized plots of the elements of the discrete Bloch
sphere, the irreducible single-qubit (two-dimensional) state vectors
with unit norm over the field $\Fpp$. We show the results for $p=3$,
$7$, and $11$. For example, in $\Fppx{3}$, there are 24 vectors
of unit norm, but only the 6 inequivalent classes appear in the plot.
The $p+1=4$ equivalent vectors in each class differ only by a complex
discrete phase.}
\label{bloch.fig} 
\end{figure}

\subsection{Counting states on the $n$-qubit Bloch sphere}

We have the unique opportunity in the finite-field approach to quantum
computing to precisely identify and enumerate the physical states.
In the conventional theory, as we have seen, we employ a generalized
Hopf fibration on the normalized states to project out a circle of
phase-equivalent states, yielding the generalized Bloch sphere.

In the introduction to this section, we sketched the counting of the
irreducible single-qubit discrete states. To count the number of inequivalent
discrete states for the general $n$-qubit case with coefficients
in $\Fpp$, we first must find the set of unit-norm states, and then
determine the equivalence classes of unit-norm states under discrete
phase transformations; we can then enumerate the list of states on
the discrete generalized Bloch sphere. By executing computer searches
of these spaces, we discovered an hypothesis for a closed-form solution
for the counting of the states, and were then able to find a rigorous
inductive proof of the enumeration, which is presented in the Appendix.

This process of describing the discrete $n$-qubit irreducible states
can again be understood geometrically by following the discrete analog
of the Hopf fibration. First, we construct the discrete version of
the quadratic unit-length form that automatically annihilates the
distinction among states differing only by a discrete phase, 
\begin{eqnarray}
\hspace*{-0.5cm}\Hat{a}=(|\alpha_{i}|^{2},...,\sqrt{2}\ \mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\ \mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,
\end{eqnarray}
where 
\begin{eqnarray}
\Hat{a}\cdot\Hat{a}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
{}From \ref{appH.sec}, we know that $p+1$ elements of this discrete
$\Sphere{2\times2^{n}-1}$ structure map to the \textit{same point\/}
in $\Hat{a}$. Each set of $(p+1)$ redundant points is, geometrically
speaking, the \textit{discrete Hopf fibration circle\/} living above
each \textit{irreducible\/} point of the $n$-qubit state description.
These $p+1$ points are interpretable as the $p$ finite points plus
the single point at infinity of the projective discrete line (see,
e.g., \cite{Arnold}).

The next part of this argument is the determination of the unit-norm
states, effectively the space of allowed discrete partitions of unity;
we cannot exactly call these ``probability-conserving'' sectors
of the state coefficients since we do not have a well defined notion
of probability, but we do have a well-defined notion of partition
of unity. The tally of unit-norm states is $p^{2^{n}-1}(p^{2^{n}}-1)$
(see \ref{appU.sec}) compared to the total number $p^{2\times2^{n}}$
of possible complex integer state vectors that could be chosen. This
unit-norm state structure is the discrete analog of $\Sphere{2\times2^{n}-1}$.

Finally, we repeat the last step of the $n$-qubit continuous Hopf
fibration process for discrete $n$-qubit states, eliminating the
discrete set of $p+1$ equivalent points that map to the same point
$\Hat{a}$ on the generalized $n$-qubit Bloch sphere. Dividing the
tally $p^{2^{n}-1}(p^{2^{n}}-1)$ of unit norm states by the $p+1$
elements of each phase-equivalent discrete circle, we find 
\[
\frac{p^{2^{n}-1}(p^{2^{n}}-1)}{p+1}=p^{2^{n}-1}(p-1)\,\prod_{k=1}^{n-1}(p^{2^{k}}+1)
\]
as the total count of unique irreducible states in a discrete $n$-qubit
configuration (see \ref{appI.sec}). The resulting object is precisely
the discrete version of $\CP{D-1}$, which we might call a \textit{discrete
complex projective space\/} or $\DCP{D-1}$, where $D=2^{n}$ as
usual.

\Comment{

\subsection{Counting states on the $n$-qubit Bloch sphere}

Although we do not have a rigorous proof for the counting of discrete
unit-norm states and the number of irreducible unit-norm states in
the discrete generalized Bloch sphere for $n$-qubit states with coefficients
in $\Fpp$, we have run extensive computer searches of the state features
for a range of compatible primes $p$ and number of qubits $n$. From
these experimental investigations, we have induced a closed-form solution
for the counting that agrees for every possible combination of parameters
that can be computed in reasonable time (the computational complexity
becomes intractable very quickly).

For $n$-qubit discrete states with coefficients in $\Fpp$, our calculation
begins with an enumeration of all possible $2^{n}$-tuples of state
coefficients in $\Fpp$; then the moduli-squared of these tuples are
summed and compared to unity in $(\mbox{mod}~p)$. Those that survive
are accumulated in a list of allowed discrete partitions of unity;
we cannot exactly call these ``probability-conserving'' sectors
of the state coefficients since we do not have a well defined notion
of probability, but we do have a well-defined notion of partition
of unity. The heuristic number that results for the tally of unit-norm
states is $p^{2^{n}-1}(p^{2^{n}}-1)$, compared to the total number
$p^{2\times2^{n}}$ of possible complex integer state vectors that
could be chosen. This unit-norm state structure is now the discrete
analog of $\Sphere{2\times2^{n}-1}$.

Finally, we construct the discrete version of the quadratic unit-length
form that automatically annihilates the distinction among states differing
only by a discrete phase, 
\begin{eqnarray}
\hspace*{-0.5cm}\Hat{a}=\{|\alpha_{i}|^{2},...,\sqrt{2}\ \mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\ \mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots\}\ ,
\end{eqnarray}
where 
\begin{eqnarray}
\Hat{a}\cdot\Hat{a}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
Computing all the elements of the discrete $\Sphere{2\times2^{n}-1}$
structure that map to the \textit{same point\/} in $\Hat{a}$, we
find a $(p+1)$-fold redundancy, exactly as for the one-qubit case;
each set of $(p+1)$ redundant points is, geometrically speaking,
the \textit{discrete Hopf fibration circle\/} (see, e.g., Arnold
\cite{Arnold}) living above each \textit{irreducible\/} point of
the $n$-qubit state description.

In summary, when the $n$-qubit continuous Hopf fibration process
is repeated for the discrete analog of $n$-qubit states, we find
that that analog of a phase ``circle'' is a discrete set $p+1$
equivalent points that map to the same point $\Hat{a}$ on the generalized
$n$-qubit Bloch sphere, and therefore there are a total of $p^{2^{n}-1}(p^{2^{n}}-1)/(p+1)$
unique irreducible states in a discrete $n$-qubit configuration.
The resulting object is precisely the discrete version of $\CP{D-1}$,
which we might call a \textit{discrete complex projective space\/}
or $\DCP{D-1}$. }
